Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-06 tex/main.tex
minor
More work on proof, almost ready to finish a draft of it.

If $t$ is positive, $\conv_0$ adds the constraint $\chi$ derived from
the subterms as a conjunction. Dually, if $t$ is negative, the
function constructs a disjunction with the negated constraint. It $t$ has no
known polarity, the constrained $\chi$ is simply returned to the caller.

@@

@@

Second, we show that if $\conv( \varphi )$ is satisfied by a
$\extendsig{\Sigma}$-interpretation $\I$, then $\varphi$ is satisfied by
$\Sigma$-interpretation $\J$. Since $\varphi$ is in definitional form with
respect to the functions defined by $\DDD$, we know $\varphi$ is of the form
$\DDD \wedge \varphi_0$. 
%We first argue that $\I$ satisfies a $\Sigma$-formula
%$\psi$ that is closed under function expansion with respect to $\DDD$, and
%where $\psi$ entails $\varphi_0$.
To do so, 
we first construct a sequence of $\extendsig{\Sigma}$-literals $M_0, M_1, \ldots$ such that $\I$ satisfies each $M_i$,
and a sequence of mappings $C^{\con f}_0, C^{\con f}_1, \ldots$ from values $\fargtype{f}^\I$ 
to tuples of terms of type $\tau_1 \times \cdots \times \tau_n$ 
for each function $\con{f}$ of type $\tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in $\sfundefs{\Sigma}$.
Let $M_0$ be a set of literals over the atoms occurring in $\varphi_0$
such that $\I$ satisfies each literal in $M_0$ and $\bigwedge M_0$ entails $\conv( \varphi_0 )$.
For each $\con{f} \in \sfundefs{\Sigma}$, let $C^{\con{f}}_0$ be the empty mapping.
For each $i \geq 0$, 
let $X_i$ be the set of constraints of the form $\exists a : {\fargtype{\con{f}}}. ( \vecfarg{\con{f}}( a ) \teq \vec t )$ occurring in $M_i$
and let $L_i$ be the others in $M_i$.
For each $\con{f} \in \sfundefs{\Sigma}$, let $C^{\con{f}}_{i+1}$ be a mapping containing $C^{\con{f}}_i$ and a minimal number of entries
such that for all $\exists a : {\fargtype{\con{f}}}. ( \vecfarg{\con{f}}( a ) \teq \vec t ) \in X_i$,
there exists a $u \mapsto \vec s \in C^{\con{f}}_{i+1}$ such that $\vecfarg{\con{f}}^\I( u ) = \vec s^\I = \vec t^\I$.
Let $M_{i+1}$ be the set containing $M_i$ as well as

Let $M^\ast$ be the limit of this sequence, which we partition into sets $X^\ast$ and $L^\ast$ in the same way as above.
Let $\psi$ be $\bigwedge L^\ast$.
We know $\psi$ is a $\Sigma$-formula since \ajr{blah}
We know $\psi$ is closed under function expansion with respect to $\DDD$ since \ajr{blah}
Finally, we know $\psi$ entails $\varphi_0$ since \ajr{blah}
Since there exists a model for $\psi$ that is closed under function expansion
with respect to $\DDD$, then since $\DDD$ is admissible, by definition there
exists a model for $\psi \wedge \DDD$, which entails $\DDD \wedge \varphi_0$,
which is $\varphi$.
\qed
\end{proofsketch}

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. This formula is unsatisfiable if and only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargtype{f}$ can be
interpreted as $\typeint$, and $\farg{f}$ can be interpreted as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if
and only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x\typ{\typeint}.\;\, \lite\bigl( x \leq 0,\allowbreak\; \con{s}(x)
\teq 0,\; \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissble.
As stated
in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
model. In more detail, let $\varphi_\con{s}$ be the formula $\forallf{\con{s}} x.\;\,
\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 )
\bigr)$. The rightmost conjunct $\exists a : {\fargtype{\con{s}}}.
\farg{\con{s}}( a ) \teq \con{c}$ ensures that any model of
formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
satisfies $\varphi_{\con{s}}[\con{c}/x]$. The nested existential
$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
a similar constraint is enforced recursively: any model satisfying $\varphi_{\con{s}}[t/x]$
must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$ \ajr{revise this sentence, not quite accurate}. Given a model $\I$ of
formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $14$, it is easy to see that
due to the structure of formula~(\ref{eq:ex-after}),
$\I$ must satisfy the following formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!13} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-14 \leq 0 \wedge \con{s}( \con{c}-14 ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, 14$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a $\Sigma$-interpretation satisfying $\psi \wedge \varphi_\con{s}$,
which means there exists a $\Sigma$-interpretation satisfying formula~(\ref{eq:ex-before}).\xend
\end{example}

>>>>

If $t$ is positive, $\conv_0$ adds the constraint $\chi$ derived from
the subterms as a conjunction. Dually, if $t$ is negative, the
function constructs a disjunction with the negated constraint. It $t$ has no
polarity, the constrained $\chi$ is simply returned to the caller.

@@

\begin{lemma}\label{lem:conv}
If $\varphi$ is a formula not containing function definitions,
and $\I$ is a $\extendsig{\Sigma}$-interpretation satisfying $\conv( \varphi )$,
then $\I$ satisfies a set of literals $L \cup X$ such that
$L$ is a set of $\Sigma$-literals that entail $\varphi$,
and $X$ is the set of literals
$\{ \exists a : {\fargtype{\con{f}}}. ( \vecfarg{\con{f}}( a ) \teq \vec t ) \mid \con{f}( \vec t ) \in \terms( L ), \con{f} \in \sfundefs{\Sigma} \}$.
\end{lemma}

@@

Second, we show that if $\conv( \varphi )$ is satisfied by a
$\extendsig{\Sigma}$-interpretation $\I$, then $\varphi$ is satisfied by
$\Sigma$-interpretation $\J$. Since $\varphi$ is in definitional form with
respect to the functions defined by $\DDD$, we know $\varphi$ is of the form
$\DDD \wedge \varphi_0$. 
We argue that $\I$ satisfies a $\Sigma$-formula
$\psi$ that is closed under function expansion with respect to $\DDD$, and
where $\psi$ entails $\varphi_0$.
To construct $\psi$, we define a sequence of sets of $\extendsig{\Sigma}$-literals $M_0 \subseteq M_1 \subseteq \ldots$ 
such that each $M_i$ is 
satisfied by $\I$ and
can be partitioned into two sets $L_i$ and $X_i$,
where $L_i$ are $\Sigma$-literals,
and $X_i$ consists of literals of the form $\exists a : {\fargtype{\con{f}}}. ( \vecfarg{\con{f}}( a ) \teq \vec t )$
for each term $\con{f}( \vec t ) \in \terms( L_i )$ where $\con{f} \in \sfundefs{\Sigma}$.
Let $M_0$ be 
a set of literals over atoms occurring in $\conv{ \varphi_0 }$
such that $\I$ satisfies each literal in $M_0$ and $\bigwedge M_0$ entails $\conv( \varphi_0 )$.
By case analysis on the return values of $\conv_0$ 
it can be shown that $M_0$ can be partitioned into $L_0$ and $X_0$ in the way described above.
For each $i \geq 0$,
let $\psi'$ be $\bigwedge \{ \conv( \varphi_f[ \vec t / \vec x ] ) \mid \con{f}( \vec t ) \in \terms( L_i ), f \in \sfundefs{\Sigma} \}$,
where $\varphi_f$ denotes the function definition of $f$ in $\DDD$.
Note that $\I$ also satisfies $\psi'$ since $\I$ satisfies $X_i$.
Let $M_{i+1}$ be the union of $M_0$ and
a set of literals $M$ over atoms occurring in $\psi'$
such that $\I$ satisfies each literal in $M$ and $\bigwedge M$ entails $\psi'$,
where we again may partition $M_{i+1}$ into $L_{i+1}$ and $X_{i+1}$.
Let $M^\ast$ be the limit of this sequence, 
which we may partition into (possibly infinite) sets $X^\ast$ and $L^\ast$ in the same way as above.
Let $\psi$ be the $\Sigma$-formula $\bigwedge L^\ast$.
We know $\psi$ is closed under function expansion with respect to $\DDD$ due to its construction.
Since there exists a model for $\psi$ that is closed under function expansion
with respect to $\DDD$, then since $\DDD$ is admissible, by definition there
exists a model for $\psi \wedge \DDD$, which entails $\DDD \wedge \varphi_0$,
which is $\varphi$.
\qed
\end{proofsketch}

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. This formula is unsatisfiable if and only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargtype{f}$ can be
interpreted as $\typeint$, and $\farg{f}$ can be interpreted as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if
and only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x\typ{\typeint}.\;\, \lite\bigl( x \leq 0,\allowbreak\; \con{s}(x)
\teq 0,\; \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissble.
As stated
in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
model. In more detail, let $\varphi_\con{s}$ be the formula $\forallf{\con{s}} x.\;\,
\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 )
\bigr)$. The rightmost conjunct $\exists a : {\fargtype{\con{s}}}.
\farg{\con{s}}( a ) \teq \con{c}$ ensures that any model of
formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
satisfies $\varphi_{\con{s}}[\con{c}/x]$. The nested existential
$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
a similar constraint is enforced recursively: any model satisfying $\varphi_{\con{s}}[t/x]$
must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$ \ajr{revise this sentence, not quite accurate}. Given a model $\I$ of
formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $14$, it is easy to see that
due to the structure of formula~(\ref{eq:ex-after}),
$\I$ must satisfy the following formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!13} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-14 \leq 0 \wedge \con{s}( \con{c}-14 ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, 14$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a $\Sigma$-interpretation satisfying $\psi \wedge \varphi_\con{s}$
which entails formula~(\ref{eq:ex-before}).\xend
\end{example}
